Position: Stop Making Unscientific AGI Performance Claims

### 2024-05-29 body.tex
final camera-ready things

Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}. Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML,rudin2019stop}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.

@@

In this work, we question the practice of using outcomes from mechanistic interpretability to support AGI claims. This is not to be seen as criticism toward the underlying methodologies in isolation, but rather in the context of current publishing practices that we repeatedly challenge throughout this work. Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight in this work~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.

Another related subfield investigates the capacity of LLMs to reason causally. Here, too, there is an opportunity to over-interpret the finding of causal information as causal understanding. Recent work has shown, for example, that LLMs can indeed correctly predict causal relationships and this may have practical use cases~\citep{kiciman2023causal}. But despite the potential utility, the authors also demonstrate that this capacity can be partially explained by memorization, rather than an actual understanding of causal relationships. Similarly,~\citet{zevcevic2023causal} provide empirical evidence indicating that current LLMs ``may talk causality but are not causal''.

Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models. Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work, albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI. The authors run a user study of human-AI interaction, in which participants performed better at a given task when they (wrongly) thought they were aided by a positively described AI. This Placebo effect was found to be robust to negative descriptions of AI. Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.

@@

Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866 in \citet{epley2007seeing}). Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.

This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.

@@

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion~\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.

The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

Reflecting on the previous two sections, we make the following concrete recommendations for future research: firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).

Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry. Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship. This argument has been raised long before in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}. Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts. The open-source community sets a positive example of how scientific artifacts should be published in many different ways. Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process. Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects. Open review platforms like OpenReview are a step in the right direction but they still fall short of what we know is technically feasible. Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged. As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests. It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making strong claims.

>>>>

Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it exhibits intelligence.}. Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims. At the same time, while currently popular in literature, inherent complexity and opaqueness in technical solutions may fundamentally be unwise to pursue in high-stakes applications~\citep{rudin2019stop}.

@@

In this work, we question the practice of using outcomes from mechanistic interpretability to support AGI claims. This is not to be seen as criticism toward the underlying methodologies in isolation, but rather in the context of current publishing practices that we repeatedly challenge throughout this work. Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight here~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.

Another related subfield investigates the capacity of LLMs to reason causally. Here, too, there is an opportunity to over-interpret the finding of causal information as causal understanding. Recent work has shown, for example, that LLMs can indeed correctly predict causal relationships and this may have practical use cases~\citep{kiciman2023causal}. But despite the potential utility, the authors also demonstrate that this capacity can be partially explained by memorization, rather than an actual understanding of causal relationships. Similarly,~\citet{zevcevic2023causal} provide evidence indicating that current LLMs ``may talk causality but are not causal''.

Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models. Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work, albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI. The authors run a user study of human-AI interaction, in which participants performed better at a given task when they (wrongly) thought they were aided by a positively described AI. This placebo effect was found to be robust to negative descriptions of AI. Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.

@@

Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866 in \citet{epley2007seeing}). Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors. This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.

@@

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion~\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}. The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

Reflecting on the previous two sections, we make the following concrete recommendations for future research:

\begin{itemize}
 \item \textbf{Acknowledgement of human bias}: researchers should be mindful of, and explicit about, risks of human bias and anthropomorphization in interpreting results, which both can be done as part of the results discussion, but also in a dedicated `limitations' section.
 \item \textbf{Stronger testing}: researchers should refrain from drawing premature conclusions about AGI, unless these are based on strong hypothesis tests.
 \item \textbf{Epistemologically robust standards}: we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and publicly accountable and collaborative iterations over how we will measure them, with explicit room for independent reviewing and external auditing by the broader community.
\end{itemize}
%firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).

Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry. Our broadest and perhaps most ambitious goal is for our research community to \textbf{move away from authorship and instead embrace contributorship}. This argument has been raised long before in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}. Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts. The open-source community sets a positive example of how scientific artifacts should be published in many different ways. Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process. Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects. Open review platforms like OpenReview are a step in the right direction, but still fall short of what we know is technically feasible. Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged. As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests. It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making strong claims.

### 2024-05-29 main.tex
trimmed down abstract slightly
final camera-ready things

\bibliography{biblio}
\bibliographystyle{icml2024}

>>>>

Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft collaboration.

\section*{CRediT Author Statement}
Following the CRediT Contributor Roles Taxonomy, we list author contributions in descending order of degree of contribution: \textbf{Conceptualization} PA, AD, AB, CL; \textbf{Data curation} PA; \textbf{Formal analysis} PA; \textbf{Funding acquisition} CL; \textbf{Investigation} PA; \textbf{Literature review} AD, AB, CL; \textbf{Methodology} PA, CL; \textbf{Project administration} PA, AD, AB; \textbf{Software} PA; \textbf{Supervision} CL; \textbf{Visualization} PA; \textbf{Writing - original draft} PA, AD, AB; \textbf{Writing - review and editing} PA, AD, AB, CL.
\bibliography{biblio}
\bibliographystyle{icml2024}
